---
title: "project_predictive"
author: "Robert Hand"
date: "11/1/2022"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(epitools)
library(mice)
library(VIM)
library(glmnet)
library(e1071)
library(gbm)
library(BART)
library(randomForest)
```


Reading Train Data
```{r}
train <- read.csv("train.csv", header = T)
head(train)
```


Splitting the "Cabin" Variable into three variables: 
Deck, cabin number, and side, port or starboard. 
Splitting the passengerID number into the group the 
passenger was in and their id within the group. 
```{r}
train %>% select(Cabin) %>% group_by(Cabin) %>% count()
train <- separate(train, Cabin, c("deck","num","side"), sep = "/", remove = FALSE)

train <- separate(train,PassengerId, c("group","group_number"), sep = "_")

```


```{r}
#taking a look at these new variables
train %>% select(deck) %>% group_by(deck) %>% count()
train %>% select(num) %>% group_by(num) %>% count()
train %>% select(side) %>% group_by(side) %>% count()

train %>% select(group_number) %>% group_by(group_number) %>% count()
#not many passengers were travelling in large groups. 

table(train$Transported,train$side)
#it appears the side of port or starboard has basically no impact. 

table(train$Transported,train$deck)
#it seems some of the decks may have disproportionate numbers of passengers dying. 
#Deck T has a tiny number of passengers. 

#setting blanks to NA
train[train==""] <- NA
```


Missing Data
```{r}
pMiss <- function(x){sum(is.na(x))/length(x)*100}
apply(train,2,pMiss)

#so about 2 to 3 percent of the data is missing for each of the variables. 


aggr_plot <- aggr(train, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, labels=names(train), cex.axis=.7, gap=3, ylab=c("Histogram of missing data","Pattern"))

#setting these variables to factors
cols <- c("HomePlanet", "CryoSleep", "deck", "side","Destination","Age","VIP","Transported","group_number")

train[cols] <- lapply(train[cols], factor)  ## as.factor() could also be used

#setting these to numeric
train$num <- as.numeric(train$num)
train$Age <- as.numeric(train$Age)
train$group <- as.numeric(train$group)

#imputing the data. Not using the original cabin variable but the broken down categories of deck, num and side. 

#dropping passenger names and the old cabin variable
train <- subset(train,select = c(-Cabin,-Name))

train_imp <- mice(train,m=3,maxit=10,seed=500)
summary(train_imp)

densityplot(train_imp)

stripplot(train_imp)

imp_data1 <- complete(train_imp,1)
imp_data2 <- complete(train_imp,2)
imp_data3 <- complete(train_imp,3)

#saving one of these imputed datasets for easy access later. 
apply(imp_data1,2,pMiss)
write.csv(imp_data1,"imp_dat1.csv")

```

Reading in the saved csv file of imputed data and fixing the variables. 

```{r}
imp_data1 <- read.csv("imp_dat1.csv",header = T)
imp_data1$Age<-as.numeric(imp_data1$Age)
imp_data1$Transported<-as.factor(imp_data1$Transported)
imp_data1$VIP<-as.factor(imp_data1$VIP)
imp_data1$CryoSleep<-as.factor(imp_data1$CryoSleep)
cols <- c("HomePlanet", "deck","side", "Destination")
imp_data1[cols] <- lapply(imp_data1[cols], factor)
cols2 <- c("group","group_number","num","RoomService","FoodCourt","ShoppingMall","Spa","VRDeck")
imp_data1[cols2]<- lapply(imp_data1[cols2], as.numeric)
imp_data1 <- subset(imp_data1, select = -c(X))

```


IMPUTING THE TEST DATA. Using the same method
as used to imput the training data. 
```{r}
test_data <- read.csv("test.csv",header = T)
pMiss <- function(x){sum(is.na(x))/length(x)*100}

test_data <- separate(test_data, Cabin, c("deck","num","side"), sep = "/", remove = FALSE)
test_data <- separate(test_data,PassengerId, c("group","group_number"), sep = "_")

test_data[test_data==""] <- NA

aggr_plot <- aggr(test_data, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, labels=names(test_data), cex.axis=.7, gap=3, ylab=c("Histogram of missing data","Pattern"))

cols <- c("HomePlanet", "CryoSleep", "deck", "side","Destination","Age","VIP","group_number")


test_data[cols] <- lapply(test_data[cols], factor)  ## as.factor() could also be used

test_data$num <- as.numeric(test_data$num)
test_data$Age <- as.numeric(test_data$Age)
test_data$group <- as.numeric(test_data$group)


#imputing the data. Not using the original 
#cabin variable but the broken down categories of duck, num and side. 

#dropping passenger names
test_data <- subset(test_data,select = c(-Cabin,-Name))


test_imp <- mice(test_data,m=3,maxit=10,seed=500)
densityplot(test_imp)
stripplot(test_imp)

test_imp_data1 <- complete(test_imp,1)
apply(test_imp_data1,2,pMiss)
write.csv(test_imp_data1, "test_imp_data1.csv")

```

```{r}
test_imp_data1 <- read.csv("test_imp_data1.csv")

test_imp_data1[cols] <- lapply(test_imp_data1[cols], factor)
test_imp_data1$Age<-as.numeric(test_imp_data1$Age)
test_imp_data1$VIP<-as.factor(test_imp_data1$VIP)
test_imp_data1$CryoSleep<-as.factor(test_imp_data1$CryoSleep)
test_imp_data1 <- subset(test_imp_data1, select = -c(X))
test_imp_data1[cols2]<- lapply(test_imp_data1[cols2], as.numeric)

```



EDA
```{r}

imp_data1 %>% select(group_number) %>% group_by(group_number) %>% count() #very few in some groups
imp_data1 %>% select(group) %>% group_by(group) %>% count()
imp_data1 %>% select(CryoSleep) %>% group_by(CryoSleep) %>% count()
imp_data1 %>% select(VIP) %>% group_by(VIP) %>% count()  #very few VIP
imp_data1 %>% select(deck) %>% group_by(deck) %>% count() #very few in deck T
imp_data1 %>% select(num) %>% group_by(num) %>% count() #very few in deck T
imp_data1 %>% select(side) %>% group_by(side) %>% count() #very few in deck T
imp_data1 %>% select(Destination) %>% group_by(Destination) %>% count()
imp_data1 %>% select(RoomService) %>% group_by(RoomService) %>% count() #most of them are zeros. 
imp_data1 %>% select(ShoppingMall) %>% group_by(ShoppingMall) %>% count()
imp_data1 %>% select(Spa) %>% group_by(Spa) %>% count()
imp_data1 %>% select(VRDeck) %>% group_by(VRDeck) %>% count()
imp_data1 %>% select(Transported) %>% group_by(Transported) %>% count()

table(imp_data1$deck,imp_data1$Transported)

#imp_data1 = subset(imp_data1, select = -c(group,group_number))
#imp_data1 <- imp_data1 %>% mutate(RoomService=case_when(RoomService==0~0,TRUE~1))
#imp_data1 <- imp_data1 %>% mutate(ShoppingMall=case_when(ShoppingMall==0~0,TRUE~1))
#imp_data1 <- imp_data1 %>% mutate(Spa=case_when(Spa==0~0,TRUE~1))
#imp_data1 <- imp_data1 %>% mutate(VRDeck=case_when(VRDeck==0~0,TRUE~1))
#imp__data1 <- imp_data1 %>% filter(deck!="T")
#imp_data1 <- imp__data1


```


Starting with some penalized regression, LASSO. 
```{r}

#prepping the data
set.seed(124)
x <- model.matrix(Transported ~., imp_data1)
y <- imp_data1$Transported
lasso_preds <- NA

#creating folds
k<- 5
fold <- sample(1:k,nrow(imp_data1),replace = TRUE)

#doing k-fold cv

for (i in 1:k){
mod.cv <- cv.glmnet(x[fold!=i,], y[fold!=i],alpha=1,family="binomial")
 lasso_model <- glmnet(x[fold!=i,],y[fold!=i],lambda=mod.cv$lambda.min,alpha = 1,family = "binomial")
 lasso_preds[fold==i] <- predict.glmnet(lasso_model,newx=x[fold==i,],type = "response")
}


prop_survive <- sum(imp_data1$Transported=="True")/length(imp_data1$Transported)

lasso_probs <- exp(lasso_preds)/(1+exp(lasso_preds))

lasso_predictions <- ifelse(lasso_probs>prop_survive,"True","False")
write.csv(lasso_predictions,"lasso_predictions.csv")

table(lasso_predictions,y)
mean(lasso_predictions==y)

#so starting out at 79% accuracy. 

#now the test predictions
xtest_dat <- model.matrix(~., test_imp_data1)

test_pred_lasso <- predict.glmnet(lasso_model,newx=xtest_dat,type = "response")
lasso_probs <- exp(test_pred_lasso)/(1+exp(test_pred_lasso))
lasso_predictions_test <- ifelse(lasso_probs>prop_survive,"True","False")
write.csv(lasso_predictions_test,"lasso_predictions_test.csv")

```

```{r}
coef <- predict(lasso_model, type = "coefficients")

coef
```



I want to try an elastic net across a range of potential values for alpha. 
```{r}
alphas <- c(0,.25,.5,.75,1)
predictions <- matrix(data=NA,nrow=nrow(imp_data1),ncol=length(alphas))


for (i in 1:k){
  for (j in 1:length(alphas)){
mod.cv <- cv.glmnet(x[fold!=i,], y[fold!=i],alpha=alphas[j],family="binomial")
 lasso_model <- glmnet(x[fold!=i,],y[fold!=i],lambda=mod.cv$lambda.min,alpha = alphas[j],family = "binomial")
 predictions[fold==i,j] <- predict.glmnet(lasso_model,newx=x[fold==i,],type = "response")
  }
}


#set all to probabilities function
probs <- function(x){
  exp(x)/(1+exp(x))
}


predictions <- as.data.frame(predictions)

#calculate probabilities
pred_probs <- data.frame(apply(predictions,2,probs))

#set to classes true or false function
classes <- function(x){
  ifelse(x>0.5036236,"True","False")
}

#determine classes
pred_classes <- data.frame(apply(pred_probs,2,classes)) 

#accuracy
accuracy <- function(x){
  mean(x==y)
}

apply(pred_classes,2,accuracy)
#the results of this are showing that the best 
#result is with the lasso with alpha = 1. 

```
It looks like the LASSO outperforms them. 

Boosting trees
```{r}

df_boosting <- imp_data1
df_boosting$I_transported <- (df_boosting$Transported == "True") + 0
df_boosting = subset(df_boosting, select = -c(Transported))
boosting_preds <- NA


for (i in 1:k){
boost_mod <- gbm(I_transported ~ .,data=df_boosting[fold!=i,], distribution = "bernoulli",n.tree=5000,interaction.depth=2,cv.folds = 5)
boosting_preds[fold==i] <- predict(boost_mod, newdata = df_boosting[fold==i, ], n.trees = 5000)
}

boosting_probs <- exp(boosting_preds)/(1+exp(boosting_preds))


predicted_classes <- ifelse(boosting_probs>prop_survive,"True","False")
boosting_predictions <- as.factor(predicted_classes)

table(imp_data1$Transported,boosting_predictions)
mean(imp_data1$Transported==boosting_predictions)

boost_preds_test <- predict(boost_mod, newdata = test_imp_data1, n.trees = 5000)
boosting_probs <- exp(boost_preds_test)/(1+exp(boost_preds_test))
predicted_classes <- ifelse(boosting_probs>prop_survive,"True","False")
boosting_predictions_test <- as.factor(predicted_classes)
write.csv(boosting_predictions_test, "boosting_predictions_test.csv")

```

Save CSV of boosting trees prediction
```{r}
write.csv(boosting_predictions, "boosting_predictions.csv")
```


RANDOM FOREST
```{r}
rf_predictions <- NA

for (i in 1:k){
  rf <- randomForest(Transported~., data = imp_data1[fold!=i,],
mtry = 4, importance = TRUE, ntree = 500, distribution = "bernoulli")
rf_predictions[fold==i] <-predict(rf, newdata = imp_data1[fold==i,])
}


rf_test_preds <- predict(rf, newdata = test_imp_data1)
write.csv(rf_test_preds,"rf_test_preds.csv")

```



```{r}
switch <- function(x){
  x<-ifelse(x==1,'False','True')
}

rf_preds <- switch(rf_predictions)

rf_means <- function(x){
  mean(x==imp_data1$Transported)
}

rf_means(rf_preds)
table(imp_data1$Transported,rf_preds)

write.csv(rf_preds,"randomforest_preds.csv")

plot(rf)
varImpPlot(rf)
```

SUPPORT VECTOR MACHINES
```{r}

k <- 3
set.seed(1234)
folds <- sample(1:k, nrow(imp_data1), replace = TRUE)
yhat <- rep(NA, nrow(imp_data1))
for (i in 1:k){
a <- svm(as.factor(Transported) ~ ., data = imp_data1[folds != i,], kernel = "radial", cost = 10)
yhat[folds == i] <- as.character(predict(a, newdata  = imp_data1[folds == i,]))
}

mean(yhat == imp_data1$Transported)
table(imp_data1$Transported,yhat)

write.csv(yhat,"svm_predictions.csv")

svm_test_preds <- as.character(predict(a, newdata = test_imp_data1))
write.csv(svm_test_preds, "svm_test_preds.csv")

a$nSV
summary(a)
```

```{r}
cols <- c("HomePlanet", "deck","side", "Destination")

imp_data1[cols] <- lapply(imp_data1[cols], factor)  ## as.factor() could also be used

```


Where is there total agreement?
```{r}

boosting_predictions<-ifelse(boosting_predictions=='2',1,0)
yhat <- ifelse(yhat=="True",1,0)
rf_preds<-ifelse(rf_preds=='True',1,0)
lasso_predictions <-ifelse(lasso_predictions=="True",1,0)
pred_df <- data.frame(cbind(lasso_predictions,rf_preds,yhat, boosting_predictions))

#Gonna see where the predictions have a disagreement.
agree <- ifelse(rowSums(pred_df) %in% c(0,4), "same","diff")
agree <- data.frame(agree)
pred_df <- cbind(pred_df,agree,imp_data1$Transported)

pred_df %>% filter(agree=="diff") %>% count()

```
not too many totally agree so abandoning the idea of the leftovers. 

Okay let's do test data predictions:
```{r}
test <- read.csv("test.csv")

boosting_predictions_test<-ifelse(boosting_predictions_test=='True',1,0)
svm_test_preds <- ifelse(svm_test_preds=="True",1,0)
rf_test_preds<-ifelse(rf_test_preds=='True',1,0)
lasso_predictions_test <-ifelse(lasso_predictions_test=="True",1,0)


test_pred_df <- data.frame(cbind(boosting_predictions_test,svm_test_preds,rf_test_preds,lasso_predictions_test))

```


how about by majority vote? 
```{r}
majority_vote_predictions_test <- NA
for (i in 1:nrow(test_pred_df)){
majority_vote_predictions_test[i]<-ifelse(sum(test_pred_df[i,1:3]==1)>1,"True","False")
}

maj_vote_test_preds <- data.frame(cbind(majority_vote_predictions_test,test$PassengerId))
maj_vote_test_preds <- maj_vote_test_preds %>% rename(PassengerId=V2,Transported=majority_vote_predictions_test)
write.csv(maj_vote_test_preds,"maj_vote_titanic_predictions.csv", row.names = F)
```


Individual SVM predictions
```{r}
svm_test_preds<-ifelse(svm_test_preds==1,"True","False")
svm_preds_final <- data.frame(cbind(svm_test_preds,test$PassengerId))
svm_preds_final <- svm_preds_final %>% rename(Transported=svm_test_preds, PassengerId=V2)
write.csv(svm_preds_final, "svm_titanic_predictions_test.csv", row.names = F)
```


Individual Rf predictions
```{r}
rf_test_preds<-ifelse(rf_test_preds==1,"True","False")
rf_test_preds_final <- data.frame(cbind(rf_test_preds,test$PassengerId))
rf_test_preds_final <- rf_test_preds_final %>% rename(Transported=rf_test_preds, PassengerId=V2)
write.csv(rf_test_preds_final, "rf_predictions_final.csv", row.names = F)
```


